---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# simevent

<!-- badges: start -->
<!-- badges: end -->

The goal of `simevent` is to provide functions for the generation and analysis of complex continuous time health care data. 

## Installation

You can install the development version of simevent from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("miclukacova/simevent")
```

## Example 1: simEventData

We load the package

```{r library}
library(simevent)
```

The underlying function is called `simEventData`. The `N` argument lets the user specify number of individuals in the simulation. We can simulate observational health care data for 100 individuals with the function call

```{r}
set.seed(2112)
data <- simEventData(N = 100)
```

The simulated data looks like

```{r}
head(data)
```

From data we can for example see that Individual $1$ has a baseline covariate $L_0$ of $62.89$, does not get treatment ($A_0 = 0$) and experienced the terminal event $1$ at Time $2.9$. One can visualize the data with use of the function `plotEventData`

```{r}
plotEventData(data)
```

The plot shows the event history of the $N = 100$ individuals, the different colors of the dots illustrate the different events. The $x$-axis is the timeline. 

### Arguments of simEventData

`simEventData` is quite flexible, and does therefore take many arguments. You can read about the different arguments on the help page

```{r}
?simEventData
```

To include additional baseline covariates, the user can specify random generator functions to generate the covariates. We will in the following include to additional baseline covariates, one with a bernoulli distribution with succes parameter 0.2, and one with a normal distribution. The functions should take as input the number of draws to make. 

```{r}
func1 <- function(N) rbinom(N, 1, 0.2)
func2 <- function(N) rnorm(N)
add_cov <- list(func1, func2)
```

The `beta` argument lets the user specify the effects of processes and covariates on the intensities of the processes. We define a new beta matrix, it should have dimensions 6x4 since we have two additional baseline covariates. 

```{r example}
# Effect on event 0 
beta0 <- c(0, 0, 0, 0, 0, 0)
# Effect on event 1
beta1 <- c(1, -1, 1, -1, -1, -1)
# Effect on event 2 (A)
beta2 <- c(0, -1, 0, 0.5, 0.5, 0)
# Effect on event 3 (L)
beta3 <- c(0, 0, 1, 0, 0, 0)
beta <- cbind(beta0, beta1, beta2, beta3)
```

We can specify a user defined `at_risk` function by

```{r}
at_risk <- function(i, L, A) {
  return(c(
    1,1,                      # You are always at risk for death and censoring
    as.numeric(A[i] == 0),    # You can experience A once
    as.numeric(L[i] < 2)))    # You can experience L twice
  }
```

This `at_risk` function for example, allows for recurrent events, letting the individual experience the $L$ event twice. We simulate data from this setting, and use the default values of $\eta$, $\nu$ and have no maximal censoring time.  

```{r}
data <- simEventData(N = 100, add_cov = add_cov, beta = beta, at_risk = at_risk)
```

The additional covariates are called $L_1$ and $L_2$, and data now looks like

```{r}
head(data)
```


### Cox Regression

It could be of interest to estimate the effects of covariates on the intensities of the different counting processes. A tool for this is the Cox proportional hazards model. We start by simulating data for $N = 10^4$ individuals.

```{r}
data <- simEventData(N = 10^4, beta = beta)
```

In order to fit a Cox proportional hazards model with the `survival` package, the data needs to be transformed into the so called \textit{tstart} \textit{tstop} format, this can be done by the function `IntFormatData`. 

```{r}
data_int <- IntFormatData(data)
```

Data in the \textit{tstart tstop} format looks like

```{r}
head(data_int)
```

The data contains the same information as the original data, only now three additional columns have been added. The column *k*, indicates the number of the event, the column *tstart* indicates the start of a time interval, and the column *tstop* indicates the end of a time interval. In each time interval the covariates (and processes) influencing the intensities of the processes remain constant. Cox proportional hazards models for the death process and operation process can be fitted by the following code

```{r}
library(survival)
# Death process
survfit_death <- coxph(Surv(tstart, tstop, Delta == 1) ~ I(L0/50) + A0 + L + A, 
                       data = data_int)
# Operation process
survfit_oper <- coxph(Surv(tstart, tstop, Delta == 2) ~ I(L0/50) + A0 + L, 
                      data = data_int[A == 0])
```

In order to conduct correct inference, only data where the individual is at risk for the event in question is included in the regression. The regression results can be seen by the summary call

```{r}
survfit_death |> summary()
```

## Example 2: Survival Data

A special case of the general setting is the survival setting, one can simulate data from a survival setting with the function `simSurvData`. 

```{r}
data <- simSurvData(100)
plotEventData(data, title = "Survival Data")
```

One can again specify the effects of $A_0$ and $L_0$ on the risk of death and censoring by the `beta` argument.

```{r}
# No effect of L0 and A0 on censoring process
beta_C <- c(0,0)
# Effect of L0 and A0 on death process
beta_D <- c(1,-1)

beta <- cbind(beta_C, beta_D)
```

And specify the parameters of the Weibull intensity for the censoring and death process. 

```{r}
eta <- c(0.2, 0.2)
nu <- c(1.05, 1.05)
```

We now call the function and visualize the data

```{r}
data <- simSurvData(100, beta = beta, eta = eta, nu = nu)
plotEventData(data, title = "Survival Data")
```


## Example 3: Competing Risk Data

You can simulate data from a competing risk setting with the function `simCRdata`. The arguments `beta`, `eta`, `nu`, work in a similar maner as above.  

```{r}
data <- simCRdata(100)
plotEventData(data, title = "Competing Risk Data")
```


## Example 4: Type 2 Diabetes

The function `simT2D` simulates health care data from a setting where patients can experience $3$ different events: Censoring (0), Death (1) and Type-2-Diabetes (2). The various arguments allow for the different scenarios, and you can read about them on the help page

```{r}
?simT2D
```

Below is a function call to `simT2D` 

```{r}
data <- simT2D(N = 100,
               cens = 1,
               eta = c(0.1,0.3,0.1,0.1), 
               nu = c(1.1,1.3,1.1,1.1),
               beta_L0_L = 1, 
               beta_A0_L = -1.1, 
               beta_L_D = 1, 
               beta_L0_D = 0)

plotEventData(data, title = "T2D data")
```


## Example 5: Unobserved Covariate Setting

The `simConfounding` function was created to simulate data from a setting where we have an unobserved confounding process. You can read about the function on the help page

```{r}
?simConfounding
```

One can simulate data from the default setting by the function call

```{r}
data <- simConfounding(100)
```

And one can simulate from user specified scenarios by the function call

```{r}
data <- simConfounding(N = 100,
                       beta_L_A = 1,
                       beta_L_D = 1,
                       beta_A_D = -1,
                       beta_A_L = -0.5,
                       beta_L0_A = 1,
                       eta = rep(0.1, 4),
                       nu = rep(1.1, 4),
                       followup = 5,
                       cens = 1,
                       op = 1)
```

For example the function call above simulates from a setting with the operation/treatment event (op = 1),
where there is a censoring process (cens = 1), and where after 5 time units everybody is censored (followup = 5). We can again visulize data by the function call to `plotEventData`. 

```{r}
plotEventData(data, title = "Confounding setting")
```
