---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# simevent

<!-- badges: start -->
<!-- badges: end -->

The goal of simevent is to provide functions for the generation and analysis of complex continuous time health care data. 

## Installation

You can install the development version of simevent from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("miclukacova/simevent")
```

## Example 1: simEventData

The underlying function is called `simEventData`. One can specify various arguments

```{r example}
library(simevent)
# Number of individuals
N <- 100
# Effect on event 0 
beta0 <- c(0, 0, 0, 0)
# Effect on event 1
beta1 <- c(1, -1, 1, -1)
# Effect on event 2 (A)
beta2 <- c(0, -1, 0, 0.5)
# Effect on event 3 (L)
beta3 <- c(0, 0, 1, 0)
beta <- cbind(beta0, beta1, beta2, beta3)
```

And then call the function

```{r}
data <- simEventData(N = N, beta = beta)
```

The simulated data looks like

```{r}
head(data)
```

One can visualize the data by

```{r}
plotEventData(data)
```


## Example 2: Survival Data

You can simulate data from a survival setting with the function `simSurvData`. 

```{r}
data <- simSurvData(100)
plotEventData(data, title = "Survival Data")
```

## Example 3: Competing Risk Data

You can simulate data from a competing risk setting with the function `simCRdata`. 

```{r}
data <- simCRdata(100)
plotEventData(data, title = "Competing Risk Data")
```
