---
title: "The simevent package"
output: 
  github_document:
    toc: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

```{r, include = FALSE}
library(data.table)
library(ggplot2)
```


# simevent

<!-- badges: start -->
<!-- badges: end -->

The goal of `simevent` is to provide functions for the generation and analysis of complex continuous time health care data. The simulated data includes variables that can be interpreted as treatment decisions, disease progression, and health factors. Currently the package contains the general function `simEventData`, and 5 wrapper functions that simulate data from specific settings e.g the survival setting or the competing risk setting using the underlying `simEventData` function. The `simEventData` function simulates data from a Cox proportional hazard model with Weibull hazards. The user can specify the parameters and the effects. 
The function `simEventCox` simulates using a different approach, namely the simulation uses the hazards from a `coxph` fit. This function also enables the user to perform interventions. The package further contains functions for plotting (`plotEventData`), for formatting (`IntFormatData`), and for performing various interventions (e.g. `intEffectAlphaT2D`). 

## Installation

You can install the development version of `simevent` from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("miclukacova/simevent")
```

## Usage

The package is loaded with the command

```{r library}
library(simevent)
```

Below we go thorugh different examples of how to use the package.

### Example 1: simEventData

This is an example of simulating data using `simEventData`.`simEventData` is a function for general event history simulations. The function is quite flexible, and does therefore take many arguments. You can read about the different arguments on the help page

```{r eval=FALSE}
?simEventData
```

The number of events simulated is determined by the length of the `eta` vector, `nu` vector or the number of columns in the `beta` matrix. The `beta` argument lets the user specify the effects of processes and covariates on the intensities of the processes. Below we define a new beta matrix. It has dimensions 9x5 since we wish to simulate from a setting with two baseline covariates, 5 processes and 2 additional baseline covariates. 

```{r}
set.seed(736)
beta <- matrix(rnorm(9*5), ncol = 5, nrow = 9)
```

To include additional baseline covariates, the user can specify random generator functions to generate the covariates. We will in the following include two additional baseline covariates, one with a Bernoulli distribution with success parameter 0.2, and one with a normal distribution. The functions should take as input the number of draws to make. We choose to specify two additional covariates by

```{r}
# Additional covariates
func1 <- function(N) rbinom(N, 1, 0.2)
func2 <- function(N) rnorm(N)
add_cov <- list("Z1" = func1, "Z2" = func2)
```

One can specify an at_risk function
```{r}
# at risk function
at_risk <- function(events) {
  return(c(
    1,1,                         # Always at risk for event 0 and 1
    as.numeric(events[3] < 2),   # Can experience event 2 twice
    as.numeric(events[4] < 1),   # Can experience event 3 once
    as.numeric(events[5] < 2)))  # Can experience event 4 twice
  }
```

This `at_risk` function for example, allows for recurrent events, letting the individual event 2 and 4 twice. We use the default values of `eta`, `nu` and have no maximal censoring time (`max_cens` = infinity). The `N` argument lets the user specify number of individuals in the simulation. We simulate data by the function call

```{r}
set.seed(973)
data <- simEventData(N = 5000, beta = beta, add_cov = add_cov, at_risk = at_risk)
```

The simulated data looks like
```{r}
head(data)
```

From data we can for example see that Individual $1$ has a baseline covariate $L_0$ of $0.816906$, does get baseline treatment ($A_0 = 1$) and experienced the event $2$ at Time $0.8638205$. 

It could be of interest to estimate the effects of covariates on the intensities of the different counting processes. A tool for this is the Cox proportional hazards model. In order to fit a Cox proportional hazards model with the `survival` package, the data needs to be transformed into the so called *tstart* *tstop* format, this can be done by the function `IntFormatData`, where you need to specify the indices of the columns that contain data regarding counting processes. 

```{r}
data_int <- IntFormatData(data, N_cols = 8:12)
```

Data in the *tstart* *tstop* format looks like

```{r}
head(data_int)
```

The data contains the same information as the original data, only now three additional columns have been added. The column *k*, indicates the number of the event, the column *tstart* indicates the start of a time interval, and the column *tstop* indicates the end of a time interval. In each time interval the covariates (and processes) influencing the intensities of the processes remain constant. Cox proportional hazards models for the $N_0$ process and $N_4$ process can be fitted by the following code

```{r}
library(survival)
# Process 0
survfit0 <- coxph(Surv(tstart, tstop, Delta == 0) ~ L0 + A0 + Z1 + Z2 + N2 + N3 + N4, 
                  data = data_int)

# Process 4
survfit4 <- coxph(Surv(tstart, tstop, Delta == 4) ~ L0 + A0 + Z1 + Z2 + N2 + N3 + N4, 
                  data = data_int[N4 < 2])
```

In order to conduct correct inference, only data where the individual is at risk for the event in question is included in the regression. The regression results can be seen by the summary call, as usual. For a sanity check, we can visualize the estimated coefficients and the true values. 

```{r}
CIs <- cbind("Par" = c(paste0(rownames(confint(survfit0)), "_N0"), 
                       paste0(rownames(confint(survfit0)), "_N4")), 
             rbind(confint(survfit0), confint(survfit4)))

rownames(CIs) <- NULL
colnames(CIs) <- c("Par", "Lower", "Upper")
CIs <- data.table(CIs)
CIs[, True_val := c(beta[-c(5,6),1],
                    beta[-c(5,6),5])]

CIs$Lower <- as.numeric(CIs$Lower)
CIs$Upper <- as.numeric(CIs$Upper)

pp <- ggplot(data = CIs)+
  geom_point(aes(y = Par, x = Lower))+
  geom_point(aes(y = Par, x = Upper))+
  geom_point(aes(y = Par, x = True_val, col = "Estimate"))+
  scale_color_manual(values = c("Estimate" = "red"))+
  geom_segment(aes(y = Par, yend = Par, x = Lower, xend = Upper))+
  xlab("Estimates")
pp
```


One can visualize the data with use of the function `plotEventData`.
```{r}
plotEventData(data[1:100,])
```
The plot shows the event history of the 100 first rows of data, the different colors of the dots illustrate the different events. The $x$-axis is the timeline. 

### The argument override_beta

We can specify an entry in the beta matrix with use of the argument `override_beta`. For example the following code sets the effect of L0 on the processes N0 and N1 equal to 0. 

```{r}
data <- simEventData(N = 1000, beta = beta, add_cov = add_cov, at_risk = at_risk,
                     override_beta = list("L0" = c("N0" = 0, "N1" = 0)))
```

In the default case the effects of the counting processes on the intensities are linear. This can be altered by the argument `override_beta`. For example the following code the effect of N2 on the process N1 is set to $0$ if $N2$ is larger than $1$. 

```{r}
data <- simEventData(N = 1000, beta = beta, add_cov = add_cov, at_risk = at_risk,
                     override_beta = list("N2 > 1" = c("N1" = 2)))
```

## Example 2: Survival Data

A special case of the general setting is the survival setting, one can simulate data from a survival setting with the function `simSurvData`. 

```{r}
data <- simSurvData(100)
plotEventData(data, title = "Survival Data")
```

One can again specify the effects of $A_0$ and $L_0$ on the risk of death and censoring by the `beta` argument.

```{r}
# No effect of L0 and A0 on censoring process
beta_C <- c(0,0)
# Effect of L0 and A0 on death process
beta_D <- c(1,-1)

beta <- cbind(beta_C, beta_D)
```

And specify the parameters of the Weibull intensity for the censoring and death process. 

```{r}
eta <- c(0.2, 0.2)
nu <- c(1.05, 1.05)
```

We now call the function and visualize the data

```{r}
data <- simSurvData(100, beta = beta, eta = eta, nu = nu)
plotEventData(data, title = "Survival Data")
```


## Example 3: Competing Risk Data

You can simulate data from a competing risk setting with the function `simCRdata`. The arguments `beta`, `eta`, `nu`, work in a similar maner as above.  

```{r}
data <- simCRdata(100)
plotEventData(data, title = "Competing Risk Data")
```


## Example 4: Type 2 Diabetes

The function `simT2D` simulates health care data from a setting where patients can experience $3$ different events: Censoring (0), Death (1) and Type-2-Diabetes (2). The various arguments allow for the different scenarios, and you can read about them on the help page

```{r eval=FALSE}
?simT2D
```

Below is a function call to `simT2D` 

```{r}
data <- simT2D(N = 100,
               cens = 1,
               eta = c(0.1,0.3,0.1), 
               nu = c(1.1,1.3,1.1),
               beta_L0_L = 1, 
               beta_A0_L = -1.1, 
               beta_L_D = 1, 
               beta_L0_D = 0)

plotEventData(data, title = "T2D data")
```


## Example 5: Unobserved Covariate Setting

The `simConfounding` function was created to simulate data from a setting where we have an unobserved confounding process. You can read about the function on the help page

```{r eval=FALSE}
?simConfounding
```

One can simulate data from the default setting by the function call

```{r}
data <- simConfounding(100)
```

And one can simulate from user specified scenarios by the function call

```{r}
data <- simConfounding(N = 100,
                       beta_L_A = 1,
                       beta_L_D = 1,
                       beta_A_D = -1,
                       beta_A_L = -0.5,
                       beta_L0_A = 1,
                       eta = rep(0.1, 4),
                       nu = rep(1.1, 4),
                       followup = 5,
                       cens = 1,
                       op = 1)
```

For example the function call above simulates from a setting with the operation/treatment event (op = 1),
where there is a censoring process (cens = 1), and where after 5 time units everybody is censored (followup = 5). We can again visulize data by the function call to `plotEventData`. 

```{r}
plotEventData(data, title = "Confounding setting")
```


## Example 5: Time Varying Effects

You can simulate from a setting with time varying effects with the function ...

```{r}
eta <- rep(0.1, 4)
term_deltas <- c(0,1)
time_var_eff <- matrix(0.3, ncol = 4, nrow = 6)
beta <- matrix(nrow = 6, ncol = 4, rnorm(4*6, sd = 0.5))
t_prime <- 1

data <- simEventDataTry(N = 3*10^4, t_prime = t_prime, time_var_eff = time_var_eff, eta = eta,
                        term_deltas = term_deltas, beta = beta, lower = 10^(-15), upper = 10^2,
                        max_events = 5)

data <- IntFormatData(data)

rows_to_split <- data[tstart <= t_prime & tstop > t_prime]
data1 <- copy(rows_to_split)[, `:=` (tstop = t_prime, Time = t_prime, Delta = -1)]
data2 <- copy(rows_to_split)[, `:=` (tstart = t_prime)]

data_new <- data[!(tstart <= t_prime & tstop > t_prime)]
data_new <- rbind(data_new, data1, data2)

setorder(data_new, ID, tstart)

data_new[, `:=` (A0_prime = A0 * (Time > t_prime), L0_prime = L0 * (Time > t_prime),
                 N2_prime = N2 * (Time > t_prime), N3_prime = N3 * (Time > t_prime))]

survfit0 <- coxph(Surv(tstart, tstop, Delta == 0) ~ L0 + A0 + N2 + N3 + A0_prime 
                  + L0_prime+ N2_prime + N3_prime, data = data_new)
survfit1 <- coxph(Surv(tstart, tstop, Delta == 1) ~ L0 + A0 + N2 + N3 + A0_prime 
                  + L0_prime+ N2_prime + N3_prime, data = data_new)
survfit2 <- coxph(Surv(tstart, tstop, Delta == 2) ~ L0 + A0 + N2 + N3 + A0_prime 
                  + L0_prime+ N2_prime + N3_prime, data = data_new)
survfit3 <- coxph(Surv(tstart, tstop, Delta == 2) ~ L0 + A0 + N2 + N3 + A0_prime 
                  + L0_prime+ N2_prime + N3_prime, data = data_new)


survfit0$coefficients ;beta[c(1,2,5,6),1]
```



