---
title: "The simevent package"
output: 
  github_document:
    toc: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

```{r, include = FALSE}
library(data.table)
library(ggplot2)
```

# simevent

<!-- badges: start -->
<!-- badges: end -->

The `simevent` package provides tools for simulating and analyzing complex continuous-time health care data.The simulated data includes variables that can be interpreted as treatment decisions, disease progression, and health factors. 

One of the core functions is the flexible function `simEventData`, which simulates from a Cox proportional hazards model with Weibull hazards. Users can specify parameters and covariate effects to create custom scenarios. In addition, the package offers several wrapper functions for common settings (e.g., survival data, competing risks) built on top of `simEventData`. The function `simEventTV` is an adaptation of the function `simEventData`. It allows for time varying effects. This feature results in a slightly slower simulation procedure, so when possible `simEventData` is recommended. 

Another branch of the package concerns simulating data from data. For this the package has two functions, respectively `simEventCox` and `simEventRF`. The first function takes as input Cox regressions and simulates data using these, while the latter takes as input a fitted random forest. 

Once data is simulated the package includes functions for plotting (`plotEventData`) and formatting (`IntFormatData`). 

One can perform interventions using the function `alphaSimDisease` and calculate the effect of interventions using `intEffectAlphaDisease`.

Below is a table attempting to provide an overview of the package functions. Each function in the package is described and 

**Functions relating to `simEventData`**

| Function | Description | Example | Example |
|-----------|--------------|----------------|----------|
| `simEventData()` | Simulates event history data in the most general setting using Cox proportional hazard model and Weibull hazards. | `N`, `beta`, `eta`, `nu` | `simEventData(N = 5000)` |
| `simSurvData()` | Simulates survival data. | `N`, `beta`, `eta`, `nu`, `cens`  | `simSurvData(N)` |
| `simCRdata()` | Simulates data from a competing risk setting with two competing causes. | `N`, `beta`, `eta`, `nu`, `cens` | `simCRdata(N)` |
| `simDisease()` | The function simulates health care data from a setting where patients can experience different events: Censoring, Death, Disease. The effects of the various process on one another can be specified by the arguments `beta_X_Y`. | `N`, `beta_L0_L` | `simDisease(N = 100, beta_L0_L = 1)` |
| `simTreatment()` | Simulates event history data with four types of events representing censoring (0), death (1), treatment (2), and covariate change (3). The effects of the various process on one another can be specified by the arguments `beta_X_Y`. | `N`, `beta_L_A` | `simTreatment(N = 100, beta_L_A = 1)` |
| `simDropIn()` | Simulates data corresponding to N individuals that are at risk for 4 or 5 events. Censoring (C), Death (D), Drop In Initiation (Z), Change in Covariate Process (L) and optionally Treatment (A). | `N`, `beta_L_A` | `simDropIn(N = 100, beta_L_A = 1)` |
| `simEventTV()` | Simulates general event history data in a setting with time-varying effects.| `N`, `beta`, `tv_eff`, `t_prime` | `simTreatment(N = 100, beta, tv_eff, t_prime)` |

**Functions for treating simulated data**

| Function | Description | Key Arguments | Example |
|-----------|--------------|----------------|----------|
| `IntFormatData()` | Transforms data into a so called tstart-tstop format, used when fitting Cox proportional hazards using the `survival` package. | `data`, `N_cols` | `IntFormatData(data, N_cols = 8:12)` |
| `plotEventData()` | Plots the event histories. | `data`, `title` | `plotEventData(data)` |

**Functions relating to simulating from data**

| Function | Description | Key Arguments | Example |
|-----------|--------------|----------------|----------|
| `simEventCox()` | Simulates new data using fitted Cox proportional hazard models | `N`, `cox_fits`, `L0_old`, `A0_old` | `simEventCox(N, cox_fits, L0_old, A0_old)` |
| `simEventRF()` | Simulates new data using a random forest fitted using the `randomForestSRC` package | `N`, `RF_fit` | `plot_summary(df)` |

**Functions for performing interventions**

| Function | Description | Key Arguments | Example |
|-----------|--------------|----------------|----------|
| `alphaSimDisease()` | Simulation and estimation of event history data from disease setting with shape parameter of disease process multiplied by alpha. | `N`, `alpha`, `tau`, `years_lost`| `alphaSimDisease(N = 10^4, alpha = 0.5, tau = 5, years_lost = FALSE)` |
| `alphaSimDropIn()` | Simulation and estimation of event history data from Drop In setting with shape parameter of Drop In process multiplied by alpha. |  `N`, `alpha`, `tau`, `years_lost` | `alphaSimDropIn(N = 10, alpha = 0.5, tau = 5, years_lost = F)` |
| `alphaSimTreat()` | Simulation and estimation in treatment setting with modified shape parameter of treatment Process | `N`, `alpha`, `tau`, `years_lost` | `alphaSimTreat(N = 10, alpha = 0.5, tau = 5, years_lost = F)` |
| `intEffectAlphaDisease()` | Simulates data from the disease setting in two scenarios. Under intervention on the shape parameter \eqn{\eta} of the disease process is multiplied by \code{alpha}, and a baseline (non-intervened) scenario.It computes the proportion of individuals who experience death or disease by a specified time \eqn{\tau} in the group \code{A0 = a0}, optionally returning years_lost. |  `N`, `alpha`, `tau`, `years_lost` | `intEffectAlphaDisease(N = 10, alpha = 0.5, tau = 5, years_lost = F)` |
| `intEffectAlphaDropIn()` | Does the same as above, just in the Drop In setting. | `N`, `alpha`, `tau` | `intEffectAlphaDropIn(N = 10, alpha = 0.5, tau = 5)` |
| `intEffectAlphaTreat()` | Does the same as above, just in the Treatment setting. | `N`, `alpha`, `tau`, `years_lost` | `intEffectAlphaTreat(N = 10, alpha = 0.5, tau = 5, years_lost = F)` |

## Installation

You can install the development version of `simevent` from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("miclukacova/simevent")
```

## Usage

The package is loaded with the command

```{r library}
library(simevent)
```

## Example 1: General Event History Simulation with `simEventData`

This is an example of simulating data using `simEventData`.`simEventData` is a function for general event history simulations. The function is quite flexible, and does therefore take many arguments. You can read about the different arguments on the help page

```{r eval=FALSE}
?simEventData
```

The number of simulated events corresponds to the length of the vectors \code{eta}, \code{nu}, or the number of columns in \code{beta}.

Below, we create a 9x5 beta matrix for 5 event processes and 4 baseline covariates:

```{r}
set.seed(736)
beta <- matrix(rnorm(9*5), ncol = 5, nrow = 9)
```

You can add baseline covariates by supplying generator functions:

```{r}
func1 <- function(N) rbinom(N, 1, 0.2)
func2 <- function(N) rnorm(N)
add_cov <- list("Z1" = func1, "Z2" = func2)
```

Define an \code{at_risk} function to specify risk windows for recurrent events:

```{r}
# at risk function
at_risk <- function(events) {
  return(c(
    1,1,                         # Always at risk for event 0 and 1
    as.numeric(events[3] < 2),   # Event 2 can occur twice
    as.numeric(events[4] < 1),   # Event 3 can occur once
    as.numeric(events[5] < 2)))  # Event 4 can occur twice
}
```

Optionally one can define an \code{at_risk_cov} function to specify risk windows as a function of covariates:

```{r}
# at risk function
at_risk_cov <- function(covariates) {
  return(c(
    1,1,                               # The covariates do not change the risk of event 0 and 1
    as.numeric(covariates[1] < 0.5),   # Only at risk of event 2 if L0 < 0.5
    as.numeric(covariates[2] == 1),    # Only at risk of event 3 if A0 = 1
    1))                                # The covariates do not change the risk of event 4
}
```

Simulate data for 5000 individuals:

```{r}
set.seed(973)
data <- simEventData(N = 5000, beta = beta, add_cov = add_cov, at_risk = at_risk,
                     at_risk_cov = at_risk_cov)
```

Preview the simulated data:

```{r}
head(data)
```

### Formatting Data for Cox Regression

Transform the data into *tstart-tstop* format with `IntFormatData` (specify columns with counting process data):

```{r}
data_int <- IntFormatData(data, N_cols = 8:12)
head(data_int)
```

Fit Cox models for processes 0 and 4:

```{r}
library(survival)

# Process 0
survfit0 <- coxph(Surv(tstart, tstop, Delta == 0) ~ L0 + A0 + Z1 + Z2 + N2 + N3 + N4, 
                  data = data_int)

# Process 4
survfit4 <- coxph(Surv(tstart, tstop, Delta == 4) ~ L0 + Z1 + Z2 + N2 + N3 + N4, 
                  data = data_int[(N4 < 2) & (A0 == 1)])
```

Visualize estimated coefficients alongside true values:

```{r}
CIs <- cbind("Par" = c(paste0(rownames(confint(survfit0)), "_N0"), 
                       paste0(rownames(confint(survfit4)), "_N4")), 
             rbind(confint(survfit0), confint(survfit4)))

rownames(CIs) <- NULL
colnames(CIs) <- c("Par", "Lower", "Upper")
CIs <- data.table(CIs)
CIs[, True_val := c(beta[-c(5,6),1],
                    beta[-c(2,5,6),5])]

CIs$Lower <- as.numeric(CIs$Lower)
CIs$Upper <- as.numeric(CIs$Upper)

pp <- ggplot(data = CIs)+
  geom_point(aes(y = Par, x = Lower))+
  geom_point(aes(y = Par, x = Upper))+
  geom_point(aes(y = Par, x = True_val, col = "Estimate"))+
  scale_color_manual(values = c("Estimate" = "red"))+
  geom_segment(aes(y = Par, yend = Par, x = Lower, xend = Upper))+
  xlab("Estimates")
pp
```

### Visualize Event Histories

Plot the first 100 individuals’ event histories:

```{r}
plotEventData(data[1:100,])
```

### Using override_beta

You can override specific entries in the beta matrix. For example, set the effect of \code{L0} on processes \code{N0} and \code{N1} to zero:

```{r}
data <- simEventData(N = 1000, beta = beta, add_cov = add_cov, at_risk = at_risk,
                     override_beta = list("L0" = c("N0" = 0, "N1" = 0)))
```

You can also specify non-linear effects, e.g., effect of \code{N2} on \code{N1} set to zero if \code{N2 > 1}:

```{r}
data <- simEventData(N = 1000, beta = beta, add_cov = add_cov, at_risk = at_risk,
                     override_beta = list("N2 > 1" = c("N1" = 2)))
```

Additionally the argument can be used to specify interaction effects, e.g., if we want an interaction effect of 1 between L0 and N1 on N2 we can specify the argument:

```{r}
data <- simEventData(N = 1000, beta = beta, add_cov = add_cov, at_risk = at_risk,
                     override_beta =list("L0 * N1" = c("N2" = 1)))
```

## Example 2: Survival Data with `simSurvData`

The function simSurvData allows you to simulate survival data for individuals at risk of both censoring (0) and an event (1).

By default, the function simulates survival data with no covariate effects:

```{r}
data <- simSurvData(100)
plotEventData(data, title = "Survival Data")
```

You can specify how baseline covariates \code{L0} and \code{A0} affect the risk of censoring and death through the \code{beta} matrix. For example:

```{r}
# Effects of L0 and A0 on the censoring hazard (first column)
beta_C <- c(0, 0)  

# Effects of L0 and A0 on the death hazard (second column)
beta_D <- c(1, -1)

# Combine into a 2x2 matrix where columns represent censoring and death hazards
beta <- cbind(beta_C, beta_D)
```

You can also set the Weibull shape (\eqn{\eta}) and scale (\eqn{\nu}) parameters for both censoring and death hazards:

```{r}
eta <- c(0.2, 0.2)
nu <- c(1.05, 1.05)
```

Use the specified parameters to simulate data and visualize it:

```{r}
data <- simSurvData(100, beta = beta, eta = eta, nu = nu)
plotEventData(data, title = "Survival Data")
```


## Example 3: Competing Risk Data with `simCRdata`

Simulate competing risk data (similar parameters as above):

```{r}
data <- simCRdata(100)
plotEventData(data, title = "Competing Risk Data")
```


## Example 4: Health Care Data with `simDisease`

The function `simDisease` simulates health care data from a setting where patients can experience $3$ different events: 

* **Censoring** (coded as 0)
* **Death** (coded as 1), and
* **Disease** (coded as 2).

You can customize the simulation scenarios by adjusting the function arguments. For detailed information about the parameters, see the help page.

Example simulation

```{r}
data <- simDisease(N = 100,
               cens = 1,
               eta = c(0.1,0.3,0.1), 
               nu = c(1.1,1.3,1.1),
               beta_L0_L = 1, 
               beta_A0_L = -1.1, 
               beta_L_D = 1, 
               beta_L0_D = 0)

plotEventData(data, title = "Disease data")
```


## Example 5: Health Care Data with `simTreatment`

Simulate data with a covariate process and treatment process. Default simulation:

```{r}
data <- simTreatment(100)
```

Custom scenario:

```{r}
data <- simTreatment(N = 100,
                       beta_L_A = 1,
                       beta_L_D = 1,
                       beta_A_D = -1,
                       beta_A_L = -0.5,
                       beta_L0_A = 1,
                       eta = rep(0.1, 4),
                       nu = rep(1.1, 4),
                       followup = 5,
                       cens = 1,
                       op = 1)
```

Plot

```{r}
plotEventData(data, title = "Treatment setting")
```


## Example 6: Time Varying Effects with `simEventTV`

One can simulate a setting with time-varying effects with the function `simEventTV`. After the time \code{t_prime} an additional effect of \code{tv_eff} is added to the beta matrix. 

```{r}
set.seed(6258)
eta <- rep(0.1, 4)
term_deltas <- c(0,1)
tv_eff <- matrix(0.5, ncol = 4, nrow = 6)
beta <- matrix(nrow = 6, ncol = 4, 0.5)
t_prime <- 1

data <- simEventTV(N = 100, t_prime = t_prime, tv_eff = tv_eff, eta = eta,
                   term_deltas = term_deltas, beta = beta, lower = 10^(-15), upper = 10^2,
                   max_events = 5)

plotEventData(data)
```

## Example 7: Simulating data from data

Imagine we have observed data:

```{r}
set.seed(373)
beta = matrix(c(0.5,-1,-0.5,0.5,0,0.5), ncol = 3, nrow = 2)
data <- simCRdata(N = 1000, beta = beta)
```

### Using `simEventCox`

If we want to use `simEventCox` for simulating new data from the observed data, we first need to fit Cox models:

```{r}
cox1 <- coxph(Surv(Time, Delta == 1) ~ L0 + A0, data = data)
cox2 <- coxph(Surv(Time, Delta == 2) ~ L0 + A0, data = data)
```

Then we can call the function `simEventCox` providing the fitted Cox models as arguments:

```{r}
cox_fits <- list("Process 1" = cox1, "Process 2" = cox2)
new_data <- simEventCox(100, cox_fits, L0_old = data$L0, A0_old = data$A0)
```

The new simulated data looks like:

```{r}
head(new_data)
```


### Using `simEventRF`

If we want to use `simEventRF` for simulating new data from the observed data, we need to fit a random forest model using `rfsrc` from the `randomForestSRC` package:

```{r}
RF_fit <- randomForestSRC::rfsrc(Surv(Time, Delta) ~ L0 + A0, data = data)
```

Then we can call the function `simEventRF` providing the fitted Random Forest as argument:

```{r}
new_data <- simEventRF(100, RF_fit, L0_old = data$L0, A0_old = data$A0, term_events = c(1,2))
```

The new simulated data looks like:

```{r}
head(new_data)
```

## Example 8: Interventions

Say you want to perform an intervention in the disease setting by scaling the intensity of one of your counting processes. Data without intervention can be simulated by scaling with the factor 1:

```{r}
alphaSimDisease(N = 10^3, alpha = 1,  tau = 5)
```

By default the function returns the proportion of individuals who experience death and the proportion of individuals who experience disease by a specified time \eqn{\tau} in group \code{A0 = a0}. The user can specify what gets returned by the function call with use of the arguments `years_lost` and `return_data`. If the  former is set to TRUE the function returns number of years lost before \eqn{\tau} of death and disease. If the latter is set to TRUE the function call returns the simulated data. 

To compare a non intervened scenario with a non intervened scenario, the function `intEffectAlphaTreat` can be used. This function simulates data from the disease setting in two scenarios. Under the intervention on the shape parameter \eqn{\eta} of the disease process, which is multiplied by \code{alpha}, and a baseline (non-intervened) scenario. It computes the proportion of individuals who experience death or disease by a specified time \eqn{\tau} in the group \code{A0 = a0}, optionally returning years lost. The function can also plot a sample of the event data for each scenario for comparison.

```{r}
intEffectAlphaDisease(N = 1000, alpha = 0.7, tau = 5, years_lost = TRUE, a0 = 1, plot = TRUE)
```

