---
title: "The simevent package"
output: 
  github_document:
    toc: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

```{r, include = FALSE}
library(data.table)
library(ggplot2)
```

# simevent

<!-- badges: start -->
<!-- badges: end -->

The `simevent` package provides tools for simulating and analyzing complex continuous-time health care data.The simulated data includes variables that can be interpreted as treatment decisions, disease progression, and health factors. 

At its core is the flexible function `simEventData`, which simulates from a Cox proportional hazards model with Weibull hazards. Users can specify parameters and covariate effects to create custom scenarios. 

In addition, the package offers several wrapper functions for common settings (e.g., survival data, competing risks) built on top of `simEventData.` It also includes functions for plotting (`plotEventData`), formatting (`IntFormatData`), and simulating interventions (e.g., `intEffectAlphaDisease`).

## Installation

You can install the development version of `simevent` from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("miclukacova/simevent")
```

## Usage

The package is loaded with the command

```{r library}
library(simevent)
```

### Example 1: General Event History Simulation with `simEventData`

This is an example of simulating data using `simEventData`.`simEventData` is a function for general event history simulations. The function is quite flexible, and does therefore take many arguments. You can read about the different arguments on the help page

```{r eval=FALSE}
?simEventData
```

The number of simulated events corresponds to the length of the vectors \code{eta}, \code{nu}, or the number of columns in \code{beta}.

Below, we create a 9x5 beta matrix for 5 event processes and 4 baseline covariates:

```{r}
set.seed(736)
beta <- matrix(rnorm(9*5), ncol = 5, nrow = 9)
```

You can add baseline covariates by supplying generator functions:

```{r}
func1 <- function(N) rbinom(N, 1, 0.2)
func2 <- function(N) rnorm(N)
add_cov <- list("Z1" = func1, "Z2" = func2)
```

Define an \code{at_risk} function to specify risk windows for recurrent events:

```{r}
# at risk function
at_risk <- function(events) {
  return(c(
    1,1,                         # Always at risk for event 0 and 1
    as.numeric(events[3] < 2),   # Event 2 can occur twice
    as.numeric(events[4] < 1),   # Event 3 can occur once
    as.numeric(events[5] < 2)))  # Event 4 can occur twice
  }
```

Simulate data for 5000 individuals:

```{r}
set.seed(973)
data <- simEventData(N = 5000, beta = beta, add_cov = add_cov, at_risk = at_risk)
```

Preview the simulated data:

```{r}
head(data)
```

### Formatting Data for Cox Regression

Transform the data into *tstart-tstop* format with `IntFormatData` (specify columns with counting process data):

```{r}
data_int <- IntFormatData(data, N_cols = 8:12)
head(data_int)
```

Fit Cox models for processes 0 and 4:

```{r}
library(survival)

# Process 0
survfit0 <- coxph(Surv(tstart, tstop, Delta == 0) ~ L0 + A0 + Z1 + Z2 + N2 + N3 + N4, 
                  data = data_int)

# Process 4
survfit4 <- coxph(Surv(tstart, tstop, Delta == 4) ~ L0 + A0 + Z1 + Z2 + N2 + N3 + N4, 
                  data = data_int[N4 < 2])
```

Visualize estimated coefficients alongside true values:

```{r}
CIs <- cbind("Par" = c(paste0(rownames(confint(survfit0)), "_N0"), 
                       paste0(rownames(confint(survfit0)), "_N4")), 
             rbind(confint(survfit0), confint(survfit4)))

rownames(CIs) <- NULL
colnames(CIs) <- c("Par", "Lower", "Upper")
CIs <- data.table(CIs)
CIs[, True_val := c(beta[-c(5,6),1],
                    beta[-c(5,6),5])]

CIs$Lower <- as.numeric(CIs$Lower)
CIs$Upper <- as.numeric(CIs$Upper)

pp <- ggplot(data = CIs)+
  geom_point(aes(y = Par, x = Lower))+
  geom_point(aes(y = Par, x = Upper))+
  geom_point(aes(y = Par, x = True_val, col = "Estimate"))+
  scale_color_manual(values = c("Estimate" = "red"))+
  geom_segment(aes(y = Par, yend = Par, x = Lower, xend = Upper))+
  xlab("Estimates")
pp
```

### Visualize Event Histories

Plot the first 100 individuals’ event histories:

```{r}
plotEventData(data[1:100,])
```

### Using override_beta

You can override specific entries in the beta matrix. For example, set the effect of \code{L0} on processes \code{N0} and \code{N1} to zero:

```{r}
data <- simEventData(N = 1000, beta = beta, add_cov = add_cov, at_risk = at_risk,
                     override_beta = list("L0" = c("N0" = 0, "N1" = 0)))
```

You can also specify non-linear effects, e.g., effect of \code{N2} on \code{N1} set to zero if \code{N2 > 1}:

```{r}
data <- simEventData(N = 1000, beta = beta, add_cov = add_cov, at_risk = at_risk,
                     override_beta = list("N2 > 1" = c("N1" = 2)))
```

## Example 2: Survival Data with `simSurvData`

The function simSurvData allows you to simulate survival data for individuals at risk of both censoring (0) and an event (1).

By default, the function simulates survival data with no covariate effects:

```{r}
data <- simSurvData(100)
plotEventData(data, title = "Survival Data")
```

You can specify how baseline covariates \code{L0} and \code{A0} affect the risk of censoring and death through the \code{beta} matrix. For example:

```{r}
# Effects of L0 and A0 on the censoring hazard (first column)
beta_C <- c(0, 0)  

# Effects of L0 and A0 on the death hazard (second column)
beta_D <- c(1, -1)

# Combine into a 2x2 matrix where columns represent censoring and death hazards
beta <- cbind(beta_C, beta_D)
```

You can also set the Weibull shape (\eqn{\eta}) and scale (\eqn{\nu}) parameters for both censoring and death hazards:

```{r}
eta <- c(0.2, 0.2)
nu <- c(1.05, 1.05)
```

Use the specified parameters to simulate data and visualize it:

```{r}
data <- simSurvData(100, beta = beta, eta = eta, nu = nu)
plotEventData(data, title = "Survival Data")
```


## Example 3: Competing Risk Data with `simCRdata`

Simulate competing risk data (similar parameters as above):

```{r}
data <- simCRdata(100)
plotEventData(data, title = "Competing Risk Data")
```


## Example 4: Health Care Data with `simDisease`

The function `simDisease` simulates health care data from a setting where patients can experience $3$ different events: 
* **Censoring** (coded as 0)
* **Death** (coded as 1), and
* **Disease** (coded as 2).

You can customize the simulation scenarios by adjusting the function arguments. For detailed information about the parameters, see the help page:

```{r eval=FALSE}
?simDisease
```

Example simulation

```{r}
data <- simDisease(N = 100,
               cens = 1,
               eta = c(0.1,0.3,0.1), 
               nu = c(1.1,1.3,1.1),
               beta_L0_L = 1, 
               beta_A0_L = -1.1, 
               beta_L_D = 1, 
               beta_L0_D = 0)

plotEventData(data, title = "Disease data")
```


## Example 5: Health Care Data with `simTreatment`

Simulate data with a covariate process and treatment process:

```{r eval=FALSE}
?simTreatment
```

Default simulation:

```{r}
data <- simTreatment(100)
```

Custom scenario:

```{r}
data <- simTreatment(N = 100,
                       beta_L_A = 1,
                       beta_L_D = 1,
                       beta_A_D = -1,
                       beta_A_L = -0.5,
                       beta_L0_A = 1,
                       eta = rep(0.1, 4),
                       nu = rep(1.1, 4),
                       followup = 5,
                       cens = 1,
                       op = 1)
```

Plot

```{r}
plotEventData(data, title = "Treatment setting")
```


## Example 5: Time Varying Effects with `simEventTV`

One can simulate a setting with time-varying effects with the function `simEventTV`. After the time \code{t_prime} an additional effect of \code{tv_eff} is added to the beta matrix. 

```{r}
set.seed(6258)
eta <- rep(0.1, 4)
term_deltas <- c(0,1)
tv_eff <- matrix(0.5, ncol = 4, nrow = 6)
beta <- matrix(nrow = 6, ncol = 4, 0.5)
t_prime <- 1

data <- simEventTV(N = 1000, t_prime = t_prime, tv_eff = tv_eff, eta = eta,
                   term_deltas = term_deltas, beta = beta, lower = 10^(-15), upper = 10^2,
                   max_events = 5)

plotEventData(data)
```



